Distributional Trees for Circular Data

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
small touch-ups in Sections 1-3
some reformulation in Section 4.1

some more reformulation Section 4.2 with "covariate" rather than
"variable", emphsizing circular _covariate_ (as opposed to _response_),
and some details re inference


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2522 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 29b93241118b95a8e23f4593b82b10c4ea6f687d

%%%%%%%% IF DEPARTMENTS ARE LISTED AS WELL CHANGE NUMBERING!!!  %%%%%%%%%%%%%%
\institute{Department of Statistics, University of Innsbruck, Innsbruck, Austria
\and Department of Atmospheric and Cryospheric Science, University of Innsbruck, Innsbruck, Austria
\and Epidemiology, Biostatistics and Prevention Institute, University of Zurich, Zurich, Switzerland}

@@

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2\pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

@@

\subsection{Ensembles/Random forests}
A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A different strategy of combining trees is employed in distributional forests 
(Schlosser~et.~al, 2019) which can also be interpreted as adaptive local likelihood models. 
For a possibly new observation $x$ averaged nearest neighbor weights $w_i(x)$ are obtained 
by counting the number of trees in which it is assigned to the same terminal node as a 
learning observation $y_i, i \in \{1,\ldots,n\}$ divided by the total number of trees.
The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
observation $x$ by applying the weighted maximum likelihood estimator
\begin{equation}
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
In that way predictions are mainly based on similar observations rather than the 
complete learning data.

\subsection{Splits in circular variables}
In order to obtain more parsimonious and more stable trees another possible extension of the 
introduced circular distributional trees is to split circular covariates into two 
flexible circle segments by searching two split points simultaneously.
In general, the search for the best separation of the data space into two intervals is based 
on the optimization problem
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution for the corresponding subsets of the response as defined by the
intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
For a linear split variable these intervals are defined by a single split point.
By treating a circular variable such as wind direction linearly
this single split point $\nu$ divides the considered space~$[0,2\pi)$ into
$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
separation at zero. 
Moreover, further sequential splits in the circular variable are often required to obtain 
data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
where both interval borders are determined purely by the data.

The required test statistic to maximally select two split points simultaneously is readily 
available in the \textsf{R} package \texttt{coin} but would need to be further enhanced as proposed 
by Hothorn and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests introduced by Bai and Perron~(2003) can be considered. 
Hence, the splitting idea can be naturally extended to a two-point search, however, for 
an unbiased and inference based selection the corresponding tests employing a proper
asymptotic null distribution still need to be implemented.

>>>>

%%%%%%%% IF DEPARTMENTS ARE LISTED AS WELL CHANGE NUMBERING!!!  %%%%%%%%%%%%%%
\institute{Department of Statistics, Universit\"at Innsbruck, Innsbruck, Austria
\and Department of Atmospheric and Cryospheric Science, Universit\"at Innsbruck, Innsbruck, Austria
\and Epidemiology, Biostatistics and Prevention Institute, Universit\"at Z\"urich, Z\"urich, Switzerland}

@@

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2\pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta 2001, for a more detailed overview).

@@

\subsection{Ensembles and random forests}
A natural extension are ensembles or forests of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
(Schlosser~et.~al, 2019). More specificially, for each possibly new
observation $x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
that is based on how often $x$ is assigned to the same terminal node as
each learning observation $y_i, i \in \{1,\ldots,n\}$.

The parameters $\mu$ and $\kappa$ are then estimated for each (new) 
observation $x$ by weighted maximum likelihood estimator based on the
adaptive nearest neighbor weights:
\begin{equation}
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $x$ whereas $w_i(x) = 1$ would correspond to the unweighted
full-sample estimates and $w_i(x) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.

\subsection{Splits in circular covariates}

In order to obtain more parsimonious and more stable trees another possible
extension for \emph{circular covariates} (with or without a \emph{circular response})
is to consider their circular nature when searching the best split into two segments.
In general, searching the best separation of a covariate into a ``left'' and ``right''
daugther node tries to maximize the segmented log-likelihood:
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$, encompassing zero. The latter
strategy is invariant to the (often arbitrary) definition of the direction at zero.

When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
far away, a simple linear split typically suffices to capture such a split (as seen for the
lagged wind direction in Figure~\ref{schlosser:fig_tree}). If both $\nu$ and $\tau$ differ
clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
might miss such a pattern.

The required test statistic to maximally select two split points simultaneously is straightforward
to accomodate in the CTree framework by providing all binary indicators corresponding
to the splints into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
increasingly slow for larger sample sizes but it might be possible to speed up computations by
exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
MOB framework the extension is not quite as straightforward but one strategy could be to
adapt double maximum tests \`a la Bai and Perron~(2003).

Hence, the splitting idea can be naturally extended to a two-point search, however, for 
an unbiased and inference-based selection the corresponding testing strategies might need
further adaption.
